Liveness-Based Garbage Collection for Lazy Languages

### 2015-12-01 paper/paper.tex
Third submission

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$, \\
  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \ell,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for {\sc car-bang} and
  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

\subsection{Soundness results}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

@@

\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
&\makecell{ $\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S,
  \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
    \kappa], e, \cred{\sigma'}$ }   &  {\sc const}
\\
\hline
& \makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $
  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\
  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S,
    \heap)}$, \\ $\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
  \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
  \rightsquigarrow $ \\ $\rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
{\sc car-select} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
    S, \heap)}$, \\ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \ell,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$ }        &
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} \\
%% \hline

%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% {\sc car-clo}
%% \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\
  $\heap(\rho(x)) \mbox{ is } \bot$  or $\heap(\rho(y)) \mbox{ is }
  \bot$} & $\rho, S,
  \heap, (+~x~y), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc prim-1-bang} 
\\
\hline

@@

\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
 &
{\sc prim-add} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-1-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-2-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (f~\myvec{x}), \sigma) =
    (\rho, S, \heap)}$,\\ $\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
{\sc funcall} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
{\sc let} \\ 
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
    \sigma) = (\rho, S, \heap)}$, \\
  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc if-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
    \sigma) = (\rho, S, \heap)}$, \\$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$   $\rightsquigarrow
\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
\cred{\clazy\sigma}$}
&
{\sc if-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\
  $\heap(\rho(x)) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc return-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) =
    (\rho, S, \heap)}$,\\ $\heap(\rho(x))~\mbox{is in WHNF with
    value}~v$} & \makecell[t]{$\rho,
  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
  \cred{\sigma'}$} &
{\sc return-whnf}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\SRETURN~x), \cred{\sigma}$  $
  \rightsquigarrow$ \\
$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
  \cred{\sigma}$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics. The differences with the small-step
  semantics have been highlighted by shading. \label{fig:minefield-semantics}}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%

>>>>

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$, \\
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for {\sc car-bang} and
  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

\subsection{Soundness result}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

@@

\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
&\makecell{ $\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S,
  \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\ell :=
    \kappa], e, \cred{\sigma'}$ }   &  {\sc const}
\\
\hline
& \makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $
  \rho', S, \heap[\ell := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  {\sc cons} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,\\
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S,
    \heap)}$, \\ $\heap(\rho(x)) \mbox{ is } (v, d)$} & \makecell[t]{$\rho, (\rho', \ell, e,
  \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
  \rightsquigarrow $ \\ $\rho', S, \heap[\ell := v], e, \cred{\sigma'}$}      &
{\sc car-select} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
    S, \heap)}$, \\ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ }& \makecell[t]{$\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow $\\ $  \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$ }        &
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho,
    S, \heap)}$, \\$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} \\
%% \hline

%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% {\sc car-clo}
%% \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\
  $\rho(x) \mbox{ is } \bot$  or $\rho(y) \mbox{ is }
  \bot$} & $\rho, S,
  \heap, (+~x~y), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc prim-bang} 
\\
\hline

@@

\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
 &
{\sc prim-add} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-1-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (+~x~y), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-2-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (f~\myvec{x}), \sigma) =
    (\rho, S, \heap)}$,\\ $\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
{\sc funcall} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}\rangle], e, \cred{\sigma}$} &
{\sc let} \\ 
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
    \sigma) = (\rho, S, \heap)}$, \\
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc if-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2),
    \sigma) = (\rho, S, \heap)}$, \\$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$   $\rightsquigarrow
\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\pi:\SIF~\psi:x~e_1~e_2), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
\cred{\clazy\sigma}$}
&
{\sc if-clo} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc return-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) =
    (\rho, S, \heap)}$,\\ $\heap(\rho(x))~\mbox{is in WHNF with
    value}~v$} & \makecell[t]{$\rho,
  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
  \cred{\sigma'}$} &
{\sc return-whnf}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\SRETURN~x), \sigma) = (\rho, S, \heap)}$,\\$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\SRETURN~x), \cred{\sigma}$  $
  \rightsquigarrow$ \\
$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
  \cred{\sigma}$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics. The differences with the small-step
  semantics have been highlighted by shading. \label{fig:minefield-semantics}}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%
